Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-07-09 paper-new/concrete-relational-model.tex
[paper] revise sections on concrete relational model

To model term precision, we begin by equipping the denotation of every type with
an ordering relation. Since term precision is reflexive and transitive, and
since we identify terms that are equi-precise, we model value types as
partially-ordered sets and values $\Gamma \vdash V : A$ as \emph{monotone}
functions. Analogously, every error domain is now equipped with a partial
ordering for which the error element is the bottom element, and the map
$\theta_B \laterhs B \to B$ is now required to be monotone. 
%
\footnote{This requires us to define a notion of ``later'' for partially-ordered
sets $X$. The elements are simply ``later''-elements of $X$, and the ordering is
the obvious one, i.e., two later-elements are related if after one time step,
the elements they step to are related.}

We model type precision $c : A \ltdyn A'$ as a \emph{monotone relation}, i.e., a
relation $c$ that is upward-closed under the relation on $A'$ and
downward-closed under the relation on $A$. The relation on the poset $A$ is
denoted $r(A)$.
% This extends to products $c_1 \times c_2$ in the obvious way.
%
Likewise, we define a notion of \emph{computation relation} on computation types
$B$ and $B'$. A computation relation $d : B \rel B'$ is a monotone relation on
the underlying sets that preserves error and respects $\theta$. That is, $\mho_B
\mathbin{d} y$ for all $y : B'$ and if $\laterhs_t (\tilde{x}_t \mathbin{d}
\tilde{y}_t)$, then $\theta_B (\tilde{x}) \mathbin{d} \theta_{B'} (\tilde{y})$.
The functors $U$ and $F$ introduced in the previous section extend to relations,
The functor $U$ simply takes the underlying monotone relation of a computation
relation; we will give the definition of $F$ on relations later in this section.

Before moving on, we introduce the notion of \emph{square}, which specifies
a relation between two monotone functions. The definition is as follows. Let
$A_i, A_o, A_i', A_o'$ be partially-ordered sets. Let $c_i : A_i \rel A_i'$ and
$c_o : A_o \rel A_o'$ be monotone relations, and let $f : A_i \to A_o$ and $g :
A_i' \to A_o'$ be monotone functions. We say that $f \ltsq{c_i}{c_o} g$ if for
all $x : A_i$ and $y : A_i'$ with $x \binrel{c_i} y$, we have $f(x) \binrel{c_o}
g(y)$. We visualize this situation as the following square:

@@

We note that for any $c$, there is a ``vertical identity'' square $\id
\ltsq{c}{c} \id$, as can be seen by unfolding the definition. Similarly, for any
montone function $f : A_i \to A_o$, there is a ``horizontal identity'' square $f
\ltsq{r(A_i)}{r(A_o)}$ by the definition of monotone function.

The cast rules UpL/UpR/DnL/DnR specify a relationship between the semantics of
type precision and the corresponding casts (see Section \ref{sec:GTLC}). In
particular, they specify that the relations are \emph{representable} by the
casts. The UpL and UpR rules refer to upcasts, and since upcasts are pure, these
rules apply to \emph{value} relations $c$. The DnL and DnR rules refer to
downcasts, and since downcasts are effectful, these rules apply to
\emph{computation} relations $d$.

@@

The existence of these simpler squares for a relation can be stated without
reference to other relations. Thus we can require that these squares exist in
the definition of relation without running into the circularity issue we faced
with the original rules.

Finally, we model the term precision ordering $\Gamma \ltdyn \Gamma' \vdash M \ltdyn
N : A \ltdyn A'$ as an ordering relation between the monotone functions denoted
by $M$ and $N$, as shown below:

@@

Now we can define the denotation of term precision for closed terms of type
$\nat$ as follows:
%
\[ l_1 \leq l_2 := 
  \Sigma_{l_1', l_2' \in \li \mathbb{N}}. l_1 \bisim l_1' \ltls l_2' \bisim l_2. \]

@@

\subsection{Adjusting the Cast Rules}

@@

So far, we have been focusing on closed terms of base type, for which the above
techniques give a semantic interpretation of term precision. We now consider how
to extend these constructions to \emph{all} types. In particular, in order to
model higher-order data types (i.e., functions) we need to equip \emph{every}
semantic object with not only a partial ordering relation but also a
``bisimilarity'' relation that is reflexive and symmetric. We must similarly
equip every object with a structured set of delays that can be inserted to
ensure the appropriate cast rules hold at all higher-order types. For example,
the upcast for a derivation $c \ra d$ involves the downcast corresponding to $c$
and an upcast corresponding to $d$. We must be able to insert delays in a
functorial manner to mimic the structure of the casts. In the next section, we
will make the necessary definitions and describe the relevant constructions.

>>>>

To model term precision, we begin by equipping the denotation of every type with
an ordering relation. Since term precision is reflexive and transitive, and
since we identify terms that are equi-precise, we model value types as
partially-ordered sets and values $\Gamma \vdash V : A$ as \emph{monotone}
functions. Analogously, every error domain is now equipped with a partial
ordering for which the error element is the bottom element, and the map
$\theta_B \laterhs B \to B$ is now required to be monotone. 
%
\footnote{This requires us to define a notion of ``later'' for partially-ordered
sets $X$. The elements are simply ``later''-elements of $X$, and the ordering is
the obvious one, i.e., two later-elements are related if after one time step,
the elements they step to are related.}
%
Morphisms of error domains are morphisms of the underlying partially-ordered
sets that preserve the error element and $\theta$ map, as was the case in the
previous section.

\subsubsection{Modelling Type Precision}
We model type precision $c : A \ltdyn A'$ as a \emph{monotone relation}, i.e., a
relation $c$ that is upward-closed under the relation on $A'$ and
downward-closed under the relation on $A$. We denote such a relation between $A$
and $A'$ by $c : A \rel A'$. The relation on the poset $A$ is denoted $r(A)$.
% This extends to products $c_1 \times c_2$ in the obvious way.
Composition of relations on predomains is the usual relational composition
(which is truncated to be propositional).

We likewise define a notion of \emph{error domain relation} between error
domains $B$ and $B'$. An error domain relation $d : B \rel B'$ is a monotone
relation on the underlying sets that preserves error and respects $\theta$. More
formally:

\begin{definition}
  Let $B$ and $B'$ be error domains. An \emph{error domain relation} between
  $B$ and $B'$ is a monotone relation $d$ on the underlying partially-ordered sets such that:
  \begin{enumerate}
     \item (Respects error): For all $y \in B'$, we have $(\mho_B, y) \in d$.
     \item (Preserves $\theta$): For all $\tilde{x}$ in $\laterhs B$ and
     $\tilde{y} \in \laterhs B'$, if $\laterhs_t ((\tilde{x}_t, \tilde{y}_t) \in
     d)$, then $(\theta_B (\tilde{x}), \theta_{B'} (\tilde{y})) \in d$. 
  \end{enumerate}
\end{definition}

Because of a technical restriction involving the interaction between
propositional truncation and the later modality, we cannot define the
composition of error domain relations as simply the composition of their
underlying predomain relations. Instead, we define composition of error domain
relations $d$ on $B_1$ and $B_2$ and $d'$ on $B_2$ and $B_3$ to be the least
relation containing $d$ and $d'$ that is downward closed under $\le_{B_1}$,
upward-closed under $\le_{B_3}$, respects error, and preserves $\theta$.
Specifically, it is defined inductively by the following rules:
%
\begin{mathpar}
    \inferrule*[right = Comp]
    {b_1 \mathbin{d} b_2 \and b_2 \mathbin{d'} b_3}
    {b_1 \mathbin{d \relcomp d'} b_3}

    \inferrule*[right = DnClosed]
    {b_1' \le_{B_1} b_1 \and b_1 \mathbin{d \relcomp d'} b_3}
    {b_1' \mathbin{d \relcomp d'} b_3}

    \inferrule*[right = UpClosed]
    {b_1 \mathbin{d \relcomp d'} b_3 \and b_3 \le_{B_3} b_3'}
    {b_1 \mathbin{d \relcomp d'} b_3'}

    \inferrule*[right = PresErr]
    { }
    {\mho_{B_1} \mathbin{d \relcomp d'} b_3}

    \inferrule*[right = PresTheta]
    {\later_t( \tilde{b_1} \mathbin{d \relcomp d'} \tilde{b_3} ) }
    {\theta_{B_1}(\tilde{b_1}) \mathbin{d \relcomp d'} \theta_{B_3}(\tilde{b_3}) }
\end{mathpar}

% That is,
% $(\mho_B, y) \in d$ for all $y : B'$ and if $\laterhs_t ((\tilde{x}_t,
% \tilde{y}_t) \in d)$, then $(\theta_B (\tilde{x}), \theta_{B'} (\tilde{y})) \in
% d$.

The functors $U$, $F$, $\times$, and $\arr$ introduced in the previous
section extend to relations. The functor $U$ simply takes the underlying
monotone relation of an error domain relation. Given a relation $c_1 : A_1 \rel
A_1'$ and $c_2 : A_2 \rel A_2'$, the relation $c_1 \times c_2 : (A_1 \times A_2)
\rel (A_1' \times A_2')$ is defined by $((x, y), (x', y')) \in (c_1 \times c_2)$
if and only if $(x, x') \in c_1$ and $(y, y') \in c_2$. Given a relation $c : A
\rel A'$ and an error domain relation $d : B \rel B'$, the relation $c \arr d :
(A \arr B) \rel (A' \arr B')$ is the error domain relation defined by $(f, g)
\in (c \arr d)$ if and only if for all $x : A$ and $x' : A'$ with $(x , x') \in
c$ we have $(f(x), g(x')) \in Ud$. The fact that this relation preserves error
and respects $\theta$ follows from the corresponding properties for $d$.
%
We we will postpone the description of the action of $F$ on relations to later
in this section.

\subsubsection{Modelling Term Precision}\label{sec:modeling-term-precision}
To model term precision, we need the notion of \emph{square}, which specifies a
relation between two monotone functions. The definition is as follows. Let $A_i,
A_o, A_i', A_o'$ be partially-ordered sets. Let $c_i : A_i \rel A_i'$ and $c_o :
A_o \rel A_o'$ be monotone relations, and let $f : A_i \to A_o$ and $g : A_i'
\to A_o'$ be monotone functions. We say that $f \ltsq{c_i}{c_o} g$ if for all $x
: A_i$ and $y : A_i'$ with $(x, y) \in c_i$, we have $(f(x), g(y)) \in c_o$. We
visualize this situation as the following square:

@@

We note that for any $c$, there is a ``vertical identity'' square $\id
\ltsq{c}{c} \id$, as can be seen by unfolding the definition. Similarly, for any
montone function $f : A_i \to A_o$, there is a ``horizontal identity'' square $f
\ltsq{r(A_i)}{r(A_o)}$ by the definition of monotone function.
We likewise make the analogous definition of square for error domains.

Having defined squares, we now discuss the interpretation of the cast rules
UpL/UpR/DnL/DnR. These rules specify a relationship between the semantics of
type precision and the corresponding casts (see Section \ref{sec:GTLC}). In
particular, they specify that the relations are \emph{representable} by the
casts. The UpL and UpR rules refer to upcasts, and since upcasts are pure, these
rules apply to \emph{value} relations $c$. The DnL and DnR rules refer to
downcasts, and since downcasts are effectful, these rules apply to
\emph{computation} relations $d$.

@@

The existence of these simpler squares for a relation can be stated without
reference to other relations. Thus, we can require that these squares exist in
the definition of relation without running into the circularity issue we faced
with the original rules.

Finally, we use squares to model the term precision ordering $\Gamma \ltdyn
\Gamma' \vdash M \ltdyn N : A \ltdyn A'$ as an ordering relation between the
monotone functions denoted by $M$ and $N$:

@@

Now we can define the denotation of term precision for closed terms of type
$\nat$ as follows:
%
\[ l_1 \leq^{\bisim} l_2 := 
  \Sigma_{l_1', l_2' \in \li \mathbb{N}}. l_1 \bisim l_1' \ltls l_2' \bisim l_2. \]

@@

\subsection{Adjusting the Cast Rules}\label{sec:adjusting-cast-rules}

@@

The above techniques give a semantic interpretation of term precision for closed
terms of base type. We now consider how to extend these constructions to
potentially-open terms of \emph{all} types. In particular, in order to model
higher-order data types (i.e., functions) we need to equip \emph{every} semantic
object with not only a partial ordering relation but also a ``bisimilarity''
relation that is reflexive and symmetric. We must similarly equip every object
with a structured set of delays that can be inserted and manipulated to ensure
the appropriate cast rules hold at all higher-order types. For example, the
upcast for a derivation $c_i \ra c_o$ involves the downcast corresponding to
$c_i$ and the upcast corresponding to $c_o$. We must be able to insert delays in
a functorial manner to mimic the structure of the casts. In the next section, we
will make the necessary definitions and describe the relevant constructions.
